Hyperactivity of mTORC1 or mTORC2-dependent signaling causes epilepsy downstream of somatic PTEN loss

Version 1

Abstract

Abstract

Gene variants that hyperactivate PI3K-mTOR signaling in the brain lead to epilepsy and cortical malformations in humans. Some gene variants associated with these pathologies only hyperactivate mTORC1, but others, such as PTEN, PIC3CA, and AKT, hyperactivate both mTORC1- and mTORC2-dependent signaling. Previous work has established a key role for mTORC1 hyperactivity in mTORopathies, however, whether mTORC2 hyperactivity contributes is not clear. To test this, we inactivated mTORC1 and/or mTORC2 downstream of early Pten deletion in a new model of somatic Pten LOF in the cortex. Spontaneous seizures and epileptiform activity persisted despite mTORC1 or mTORC2 inactivation alone, but inactivating both mTORC1 and mTORC2 normalized pathology. These results suggest that hyperactivity of both mTORC1 and mTORC2 are sufficient to cause epilepsy, and that targeted therapies should aim to reduce activity of both complexes.

Introduction

Genetic disruptions that hyperactivate mTOR activity, termed ‘mTORopathies’, cause macrocephaly, intractable childhood epilepsies, and behavioral presentations including autism spectrum disorders [1-4]. The mTOR kinase can function in two complexes, mTORC1 and mTORC2, each with unique downstream targets and upstream activators [5]. A unifying feature of mTORopathies is hyperactivation of mTORC1, which has led to the hypothesis that this complex and its effectors mediate disease phenotypes [6]. A subset of mTORopathies however, (e.g. those caused by variants in PTEN, PIC3CA, and AKT), hyperactivate both mTORC1 and mTORC2 [7,8], and mTORC2 hyperactivity has also been reported in human TLE [9]. It is unclear whether mTORC2 hyperactivity contributes to, or can cause, disease phenotypes such as epilepsy.

PTEN LOF in neurons induces soma hypertrophy, increased dendritic branching, and synaptic hyperconnectivity [10,11]. These phenotypes resemble those observed in models of mTORC1-specific hyperactivation [12], and can be prevented with either the mTORC1 inhibitor rapamycin or loss of the mTORC1-specific protein Raptor [13]. Rapamycin also rescues epilepsy and mortality associated with Pten loss, even after epilepsy has been established, suggesting that mTORC1 hyperactivity underlies these phenotypes [14-16]. Thus, there is strong evidence that hyperactivation of mTORC1 downstream of PTEN disruption causes the macrocephaly, epilepsy, early mortality, and synaptic dysregulation observed in humans and model organisms [17].

There is also evidence against the mTORC1-centric hypothesis. Rapamycin may affect mTORC2 activity at high doses or over long periods of time [18,17], and mTORC2 regulates synapse function, neuronal size, and cytoskeletal organization independent of mTORC1 [19,20]. In a mouse model in which Pten was inactivated in forebrain neurons in early adulthood, mTORC2 inhibition, but not mTORC1 inhibition, rescued seizures and behavioral abnormalities. In this model, epilepsy was dissociated from macrocephaly, which was rescued by mTORC1 inhibition but not mTORC2 inhibition [21]. However, mTORC2 inactivation did not normalize spontaneous seizures or seizure susceptibility in a model of Pten loss in dentate granule neurons [22]. Thus, there is conflicting evidence about whether mTORC1 or mTORC2 hyperactivity causes epilepsy downstream of Pten loss.

To address this, we suppressed mTORC1 and mTORC2 activity alongside PTEN LOF by inducing simultaneous deletion of Raptor or Rictor, whose protein products are unique and essential components of mTORC1 and mTORC2, respectively, in a model of somatic PTEN LOF in the cortex. In this model, epilepsy could be rescued by concurrent mTORC1 and mTORC2 inactivation, but persisted when either gene remained intact, suggesting that hyperactivity of either complex can lead to neuronal hyperexcitability and epilepsy.

Results

Generation of a developmental brain somatic mosaic model of PTEN LOF

To model the epileptogenic somatic mutations often observed in mTORopathies [23,24], we injected an AAV9 virus expressing GFP and Cre under control of the hSyn promoter into one hemisphere of the cortex of Pten(fl/fl), Pten(fl/fl)-Raptor(fl/fl), Pten(fl/fl)-Rictor(fl/fl), and Pten(fl/fl)-Raptor(fl/fl)-Rictor(fl/fl) mice at P0 (Fig. 1A). Green fluorescent protein (GFP) was expressed in neurons throughout the cortical layers in all experimental animals, but largely confined to one cortical hemisphere and the underlying hippocampus (Fig. 1B). The percentage of cells that were GFP+ was similar across groups, although there was a decrease in cell density in the Pten LOF and Pten-Rictor LOF groups [25] (Fig. 1C).

Characterization of the focal cortical Pten LOF model.

A) Experimental timeline showing induction of hSyn-Cre-GFP or Control hSyn-GFP AAV at P0 and EEG recording in adulthood. B) Representative images of GFP expression in a Control mouse brain, demonstrating expression predominantly in one hemisphere of the cortex. C1) Quantification of lesion severity and neuron density in Cre-expressing animals. There were fewer Cre-expressing neurons per unit area in the Pten LOF and Pten-Rictor LOF groups, which was at least partially attributable to a decrease in cell density in these groups. No significant differences in Cre expression remained when Cre expression was calculated based on cell density rather than area. C2) Representative images of DAPI and Cre fluorescence in the cortex of Pten LOF and Pten-Raptor LOF animals. Cell density and Cre density in Pten-Raptor LOF animals was indistinguishable from Controls. D1) Phospho-S6, a marker of mTORC1 activity, was increased by Pten LOF and reduced to control levels by concurrent Raptor LOF. Phospho-S6 was also increased from Control levels in Pten-Rictor LOF, indicating that mTORC1 hyperactivity was not normalized by Rictor LOF. Combined Raptor/Rictor LOF also normalized phospho-S6 expression. D2) phospho-Akt, a marker of mTORC2 activity, was increased in Pten LOF and normalized by Rictor LOF, but not by Raptor LOF. Combined Raptor/Rictor LOF also normalized phospho-Akt expression. Error bars show mean ± s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, *** indicates p<0.001, and **** indicates p<0.0001 as assessed by one-way ANOVA with Tukey multiple comparisons correction. Diagram created with BioRender.com.

We analyzed the impact of Pten, Raptor, and Rictor LOF on mTORC1 and mTORC2 activity in the affected hemisphere using quantitative immunohistochemistry. Phospho-S6 levels, which report mTORC1 activity through S6K1, were increased from Controls in Pten LOF and Pten-Rictor LOF brains, but rescued in Pten-Raptor LOF brains (Fig. 1D1). Phospho-Akt (S473) levels, which report mTORC2 activity, were increased in Pten LOF and Pten-Raptor LOF but normalized in Pten-Rictor LOF brains (Fig. 1D2). Levels of both pS6 and pAkt were not different from Controls in Pten-Raptor-Rictor LOF brains. These results indicate that Pten LOF hyperactivates both mTORC1 and mTORC2 in this model, and that these increases in activity are independently rescued by mTORC1 and mTORC2 inactivation.

Raptor and/or Rictor LOF attenuate Pten LOF-induced macrocephaly

Macrocephaly and neuronal hypertrophy are well-characterized sequelae of deficient PTEN signaling in patients and mouse models [26,27]. In mouse models of Pten LOF, these phenotypes can be rescued by rapamycin treatment or Raptor LOF [21,28,17,13]. Here, cortical thickness was increased in Pten LOF and rescued to Control levels by Raptor LOF, Rictor LOF, or combined Raptor-Rictor LOF (Fig. 2A-B). In individual neurons Pten LOF caused a ≈100% increase in mean soma size versus Control. Pten-Rictor LOF significantly reduced soma size from Pten LOF levels, but still showed a 60% increase over Controls. Soma size in Pten-Raptor LOF and Pten-Raptor-Rictor LOF neurons did not significantly differ from Controls (Fig. 2C).

mTORC1 inhibition or dual mTORC1/2 inhibition rescues, and mTORC2 inhibition partially rescues, abnormalities in cortical morphology induced by focal Pten LOF.

A) Example images showing fluorescent Nissl stain (red) and GFP expression (green) in cortical neurons. The top row shows the cortical thickness in all 5 groups and the bottom row shows zoomed in images depicting the differences in soma size across groups. B) The mean cortical thickness was increased in Pten LOF throughout the cortex. Pten-Raptor, Pten-Rictor, and Pten-Raptor-Rictor LOF cortical thickness did not differ significantly from Controls. C) The mean soma size was strongly increased in Pten LOF and to a smaller extent in Pten-Rictor LOF. Pten-Raptor LOF and Pten-Raptor-Rictor LOF groups did not differ significantly from Controls. Error bars show mean ± s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, and **** indicates p<0.0001 as assessed by one-way ANOVA with Tukey multiple comparisons correction.

Concurrent mTORC1/2 inactivation, but neither alone, rescues epilepsy and interictal EEG abnormalities in focal Pten LOF

Next, we measured epileptic brain activity with video-EEG to determine the ability of mTORC1 or mTORC2 activity to rescue this key feature of mTORopathies, Generalized seizures (GS) were observed in 0/9 Control, 4/7 Pten LOF, 2/6 Pten-Raptor LOF, 2/7 Pten-Rictor LOF, and 0/6 Pten-Raptor-Rictor LOF animals. GS events were longer-lasting in Pten-Raptor LOF animals than in Pten LOF or Pten-Rictor LOF animals (Fig. 3A). GS did not appear to be correlated with mTOR pathway activity (Supplementary Figure 2).

Combined mTORC1 and mTORC2 inactivation, but neither alone, rescues epilepsy in the focal Pten LOF model.

Spontaneous seizures and interictal spike activity were assessed in mice expressing Pten LOF, Pten-Raptor LOF, Pten-Rictor LOF, Pten-Raptor-Rictor LOF, and Controls. A) Representative traces of GS in a subset of animals in Pten LOF, Pten-Raptor LOF, and Pten-Rictor LOF groups. B) Spectrograms (top) and traces depicting example SWD event trains. C) Summary data showing the number of GS per day and GS length. GS events were significantly longer in the Pten-Raptor LOF group than the other groups. In the Pten LOF group, 19% of GS events (9/48) exceeded 45 seconds in length, and these events were observed in 2/4 Pten LOF GS+ animals. 79% of GS events (15/19) in the Pten-Raptor LOF group exceeded 45 seconds, and these events were observed in 2/2 GS+ Pten-Raptor LOF animals. 1/11 GS events in the Pten-Rictor LOF group exceeded this threshold. D. Summary data showing the SWD rate and length in all animals. Error bars show mean ± s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, *** indicates p<0.001, and **** indicates p<0.0001 as assessed by one-way ANOVA with Tukey multiple comparisons correction.

The most striking and consistent type of epileptic brain activity in the Pten LOF animals was frequent 5-7 Hz spike trains lasting 3 seconds or more, which fit previous characterizations of spike- and-wave discharges (SWDs) in rodents [29]. SWDs were observed in all Pten LOF, Pten-Raptor LOF, and Pten-Rictor LOF animals, and in 3/9 Control and 3/6 Pten-Raptor-Rictor LOF animals. The frequency of SWDs in Pten LOF, Pten-Raptor LOF, and Pten-Rictor LOF animals was significantly higher than Controls, but Pten-Raptor-Rictor LOF completely blocked this increase. Pten-Rictor LOF animals had a lower frequency of SWD events than Pten LOF animals (Fig. 3B), indicating a partial rescue. Taken together, these data indicate inactivating mTORC1 provides no protection against Pten loss-induced epilepsy, and may even exacerbate it. mTORC2 inactivation provides some protection, but only concurrent mTORC1 and mTORC2 inactivation can prevent it.

In addition to epileptic activity, we also found that Pten LOF caused obvious alterations in features of the interictal EEG. To quantify these changes, we measured EEG coastline, absolute mean amplitude, and power spectra of the EEG. Coastline, mean amplitude, and total power were all significantly increased by Pten LOF, and these changes were not significantly reduced by Pten-Raptor or Pten-Rictor LOF. Pten-Raptor-Rictor LOF, however, reduced these increases to Control levels. Correspondingly, EEG power was increased in Pten LOF animals (Fig. 4A; Table 1A). This increase was normalized by Pten-Raptor-Rictor LOF, but not by Pten-Raptor LOF or Pten-Rictor LOF. Pten-Raptor-Rictor LOF did not fully rescue a rightward shift in normalized EEG band power (Fig. 4B; Table 1B).

Combined mTORC1 and mTORC2 inactivation, but neither alone, rescues Pten LOF-induced abnormalities in the interictal EEG.

A) Examples of typical EEG traces for each genotype. In EEG epochs that were not characterized as GTCS or SWD events, Pten LOF animals had higher levels of EEG activity as quantified by EEG line length, absolute mean amplitude, and total power. These changes were not significantly decreased by Raptor LOF and Rictor LOF but were normalized by combined Raptor/Rictor LOF. B) Total EEG power was increased by Pten LOF and attenuated but not normalized by either Raptor or Rictor LOF. Relative power was decreased in delta and increased higher frequencies by Pten LOF. Pten-Raptor LOF, Pten-Rictor LOF, and Pten-Raptor-Rictor LOF animals all showed a milder rightward shift of EEG power. C) Line length, mean amplitude, and power are increased in Pten LOF and normalized by Pten-Raptor-Rictor LOF. Error bars show mean ± s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, *** indicates p<0.001 as assessed by one-way ANOVA with Tukey multiple comparisons correction. Two-way ANOVA p-values for EEG power are reported in Tables 1 and 2.

Discussion

Convergent discoveries have positioned mTORopathies as prime candidates for a precision medicine approach [30,31]. There has been some clinical success with rapamycin analogues [32], but preclinical animal studies have suggested ways to further improve the selectivity and efficacy of the targets [12]. For the subset of mTORopathies in which mTORC1 but not mTORC2 signaling is hyperactive, selective inhibition of mTORC1 and its downstream effectors rescues many animal phenotypes, including seizures [33,34]. Studies using rapamycin as an mTORC1 inhibitor also came to this conclusion in Pten LOF models, but potential inhibition of mTORC2 and the finding that genetic inhibition of mTORC1 via Raptor deletion did not stop seizures, whereas inhibition of mTORC2 did [21], challenged this view. Here, we tested whether mTORC1 or mTORC2 inhibition alone were sufficient to block disease phenotypes in a model of somatic Pten LOF. Neither was, which agrees with previous findings that seizures persist in both Pten-Raptor LOF and Pten-Rictor LOF animals in separate model systems [21,22]. This data could be interpreted as suggesting that epilepsy downstream of Pten LOF can proceed via mTORC-independent mechanisms, such as elevated ß-catenin or its protein phosphatase activity [25,35]. However, we found that simultaneous mTORC1 and mTORC2 inhibition corrected all of the pathological features of our model except for a rightward shift in relative EEG power (Fig. 4B). This suggests that epilepsy can be caused by hyperactivity of either mTORC1 or mTORC2-dependent signaling downstream of Pten LOF, but not alternative pathways. Future studies should determine whether this is also true for other gene variants that hyperactivate both mTORC1 and mTORC2, as well in other models that have fully penetrant generalized seizures.

Epilepsy in Pten-Rictor LOF mice likely emerges via mechanisms similar to those in mTOR pathway variants such as TSC that hyperactivate mTORC1 but not mTORC2, although additional effects of mTORC2 inactivation may contribute [22]. Epilepsy in Pten-Raptor LOF occurs in the absence of macrocephaly, a hallmark of mTORC1 hyperactivity, possibly due to mTORC2 hyperactivity’s effect on synaptic transmission [20]. Roles for both complexes in emergent neural network function are corroborated by findings that behavior can be altered by either mTORC1 or mTORC2 inactivation [36]. Rapamycin derivatives are somewhat selective for mTORC1 [37,38] and are currently being tested in the clinic to treat mTORopathies and PTEN-related disorders [18]. Even more selective targeting of mTORC1 than can be achieved through rapamycin and its derivatives has been proposed as a way to effectively treat disease with fewer side effects [39]. Our data suggest that this may only be true for mTORopathies that only hyperactive mTORC1. For others, including PTEN, PIK3CA, and AKT, dual inhibitors or inhibitors of PI3K, which are being tested preclinically [40-42], are likely to be required.

Materials and Methods

Animals

Raptor (Jackson Labs #013188) and Rictor (Jackson Labs #020649) homozygous floxed mice were crossed to a Pten homozygous floxed line ([43] Jackson Labs #006440). Experimental animals were the offspring of two animals homozygous floxed for Pten and heterozygous floxed for Raptor and/or Rictor. Intracranial viral injections were done at P0 or P1. Animals that were homozygous for the floxed Pten allele and either homozygous or wild-type at the Raptor and/or Rictor allele were injected with an AAV9 viral vector expressing an eGFP-Cre fusion driven by the hSyn promoter (Addgene #105540). Control animals were of the same genotype but were injected with an AAV9 virus expressing eGFP under control of the hSyn promoter, but not Cre (Addgene #105539). Pup injections were conducted with hypothermic anesthesia. A Nanoject was used to deliver 100 nL of Cre or control virus to three sites spanning the left hemisphere of the cortex. Pups were quickly rewarmed and returned to their home cage. Additional control animals with no floxed genes on a C57B6/J background (Jackson Labs #000644) were injected with the eGFP-Cre expressing AAV9 to ensure that Cre expression did not have effects independent of the Ptenfl/fl genotype (Supplementary Figure 1).

Surgeries and EEG Recordings

Animals were aged to six weeks and then implanted with a wireless EEG monitoring headcap. Animals were anesthetized with 4% isofluorane and arranged on a stereotaxic surgical apparatus with ear bars. Anesthesia was maintained with 1.5-2.5% isofluorane. The skull surface was exposed and holes were made with a 23g needle at six locations on the skull surface. 3/32” screws (Antrin Miniature Specialties) were inserted into the holes. Screws were secured with VetBond then covered with surgical cement. The screws were attached to a six-pin Millimax strip which was secured to the skull with additional cement. Animals were administered 5 mg/kg ketoprofen post-surgery and allowed to recover for at least 5 days before EEG recording. Surgical protocol was based on guidance provided by Pinnacle Technologies (https://www.pinnaclet.com).

After recovery from surgery, animals were fitted with wireless three-channel EEG preamplifiers (Pinnacle Technologies) and EEG signal was recorded at 256 Hz with simultaneous video recording. EEG recording session length was dependent on battery life. A total of at least 150 hours of EEG data was collected for each experimental animal.

Histology and imaging

EEG-implanted animals were euthanized by isofluorane overdose followed by transcardiac perfusion with 4% paraformaldehyde. Brains were postfixed in 4% paraformaldehyde for 24 hours and then transferred to 30% sucrose for at least 48 hours. 40 μm coronal slices were cut and preserved in a cryoprotectant solution (50% PBS, 30% ethylene glycol, 20% glycerol). Prior to staining, slices were washed three times with PBS. Eight cross-sectional unstained slices were mounted with DAPI Fluoromount-G (SouthernBiotech) and used to assess cortical thickness, GFP expression, and cell density. For other analyses, three sections per animal at approximately Bregma -1.6, -2,1, and -2.6 were used. A fluorescent Nissl stain (Invitrogen N21482, 1:50) was used to assess soma size.

After washing, slices were placed in blocking solution (10% normal goat serum, 0.1% Triton X-100, and PBS) for 1 hour. Slices were incubated in the following primary antibodies for 3 hours at room temperature: phospho-S6 Ribosomal Protein Ser240/244 (rabbit monoclonal, 1:1000 dilution, Cell Signaling Technology, catalog #5364, RRID: AB_10694233), phospho-AKT Ser473 (rabbit monoclonal, 1:1000 dilution, Cell Signaling Technology, catalog #4060), and NeuN (guinea pig polyclonal, 1:1000 dilution, Synaptic Systems, catalog #266004). Following primary antibody application, slices were washed three times in PBS and the incubated in AlexaFluor secondary antibodies goat anti-guinea pig 647 and goat anti-rabbit 594 (Invitrogen) for 1 hour at room temperature.

Slides stained for Nissl substance, NeuN/pS6, and NeuN/pAkt were imaged at 10x (2048×2048) with 5 μm z-step on a Nikon C2 confocal microscope (UVM Microscopy Imaging Core). One image of the left (GFP-expressing) cortical hemisphere in each of three region-matched slices per animal was collected for analysis. Widefield images (2560×2160) of unstained DAPI-mounted cross-sectional slices were taken with a Zyla sCMOS camera (Andor) mounted on an upright microscope (IX73; Olympus) at 5x and 20x resolution. Pixel width was manually calculated using an 0.1 mm hemocytometer.

EEG analysis

EEG files were converted from Pinnacle’s proprietary format to European Data Format (EDF) files and imported to Matlab. A filtering program was used to flag traces in which seizures were likely to be occurring based on amplitude and line length between data points. For Control and Pten-Raptor-Rictor LOF animals, all 10 second epochs in which signal amplitude exceeded 400 μV at two time points at least 1 second apart were manually reviewed by a rater blinded to genotype. This method was not suitable for the Pten LOF, Pten-Raptor LOF, and Pten-Rictor LOF animals because of the density of high-amplitude interictal activity they displayed. For these animals, at least 100 of the highest-amplitude traces were manually reviewed and then traces with persistent abnormally low amplitude, often indicating postictal suppression, were reviewed as well. Flagged traces were displayed for a rater to mark the beginning and end of each seizure. Spike-and-wave discharge events were manually marked in 8 evenly spaced one-hour epochs in each of two 24-hour recording sessions per animal and verified on video not to be caused by movement such as chewing or scratching.

Baseline EEG measures were taken from a representative sample of EEG files for each animal. 692 evenly spaced five-second epochs were sampled over 24 hours, repeated for two recording sessions per animal. Line length was defined as the sum of the linear distances between adjacent data points during the five-second analysis epoch. Mean amplitude was defined as the mean of the absolute value of data points in the five-second analysis epoch. Power spectral density was calculated with the pwelch function in Matlab. All three EEG channels were analyzed and the mean of all channels was used for statistical analysis.

Image Analysis

Image analysis was conducted using ImageJ/Fiji. Soma size was measured by dividing Nissl stain images into a 10 mm2 grid. The somas of all GFP-expressing cells fully within three randomly selected grid squares in Layer II/III were manually traced. pS6 and pAkt expression were measured by drawing 200 μm wide columns spanning all cortical layers. Background was subtracted from images with a 30 μm rolling ball algorithm (Fiji/ImageJ). The mean pixel value of the column was recorded and values were averaged by animal.

Statistical analysis

Prism 9 (GraphPad Prism) was used to conduct statistical analyses and create graphs. Group differences were assessed using one-way ANOVA with Tukey post-hoc analysis in GraphPad9, except where noted.

eLife Assessment

This useful study addresses epilepsy caused by the loss of a molecule called Pten, resulting in hyperactivity of the mTOR pathway. The findings suggest that inhibiting two molecules called mTORC1 and mTORC2 can reduce epilepsy symptoms but there is much less effect when inhibited separately. The evidence supporting the conclusions is currently incomplete, but could be strengthened after additional experiments.

Reviewer #1 (Public Review):

Hyperactivation of mTOR signaling causes epilepsy. It has long been assumed that this occurs through overactivation of mTORC1, since treatment with the mTORC1 inhibitor rapamycin suppresses seizures in multiple animal models. However, the recent finding that genetic inhibition of mTORC1 via Raptor deletion did not stop seizures while inhibition of mTORC2 did, challenged this view (Chen et al, Nat Med, 2019). In the present study, the authors tested whether mTORC1 or mTORC2 inhibition alone was sufficient to block the disease phenotypes in a model of somatic Pten loss-of-function (a negative regulator of mTOR). They found that inactivation of either mTORC1 or mTORC2 alone normalized brain pathology but did not prevent seizures, whereas dual inactivation of mTORC1 and mTORC2 prevented seizures. As the functions of mTORC1 versus mTORC2 in epilepsy remain unclear, this study provides important insight into the roles of mTORC1 and mTORC2 in epilepsy caused by Pten loss and adds to the emerging body of evidence supporting a role for both complexes in the disease development.

Strengths:

The animal models and the experimental design employed in this study allow for a direct comparison between the effects of mTORC1, mTORC2, and mTORC1/mTORC2 inactivation (i.e., same animal background, same strategy and timing of gene inactivation, same brain region, etc.). Additionally, the conclusions on brain epileptic activity are supported by analysis of multiple EEG parameters, including seizure frequencies, sharp wave discharges, interictal spiking, and total power analyses.

Weaknesses:

The sample size of the study is small and does not allow for the assessment of whether mTORC1 or mTORC2 inactivation reduces seizure frequency or incidence. This is a limitation of the study.

The authors describe that they inactivated mTORC1 and mTORC2 in a new model of somatic Pten loss-of-function in the cortex. This is slightly misleading since Cre expression was found both in the cortex and the underlying hippocampus, as shown in Figure 1. Throughout the manuscript, they provide supporting histological data from the cortex. However, since Pten loss-of-function in the hippocampus can lead to hippocampal overgrowth and seizures, data showing the impact of the genetic rescue in the hippocampus would further strengthen the claim that neither mTORC1 nor mTORC2 inactivation prevents seizures.

Some of the methods for the EEG seizure analysis are unclear. The authors describe that for control and Pten-Raptor-Rictor LOF animals, all 10-second epochs in which signal amplitude exceeded 400 μV at two time-points at least 1 second apart were manually reviewed, whereas, for the Pten LOF, Pten-Raptor LOF, and Pten-Rictor LOF animals, at least 100 of the highest-amplitude traces were manually reviewed. Does this mean that not all flagged epochs were reviewed? This could potentially lead to missed seizures. Additionally, the inclusion of how many consecutive hours were recorded among the ~150 hours of recording per animal would help readers with the interpretation of the data.

Finally, it is surprising that mTORC2 inactivation completely rescued cortical thickness since such pathological phenotypes are thought to be conserved down the mTORC1 pathway. Additional comments on these findings in the Discussion would be interesting and useful to the readers.

Reviewer #2 (Public Review):

Summary:

The study by Cullen et al presents intriguing data regarding the contribution of mTOR complex 1 (mTORC1) versus mTORC2 or both in Pten-null-induced macrocephaly and epileptiform activity. The role of mTORC2 in mTORopathies, and in particular Pten loss-off-function (LOF)-induced pathology and seizures, is understudied and controversial. In addition, recent data provided evidence against the role of mTORC1 in PtenLOF-induced seizures. To address these controversies and the contribution of these mTOR complexes in PtenLOF-induced pathology and seizures, the authors injected a AAV9-Cre into the cortex of conditional single, double, and triple transgenic mice at postnatal day 0 to remove Pten, Pten+Raptor or Rictor, and Pten+raptor+rictor. Raptor and Rictor are essentially binding partners of mTORC1 and mTORC2, respectively. One major finding is that despite preventing mild macrocephaly and increased cell size, Raptor knockout (KO, decreased mTORC1 activity) did not prevent the occurrence of seizures and the rate of SWD event, and aggravated seizure duration. Similarly, Rictor KO (decreased mTORC2 activity) partially prevented mild macrocephaly and increased cell size but did not prevent the occurrence of seizures and did not affect seizure duration. However, Rictor KO reduced the rate of SWD events. Finally, the pathology and seizure/SWD activity were fully prevented in the double KO. These data suggest the contribution of both increased mTORC1 and mTORC2 in the pathology and epileptic activity of Pten LOF mice, emphasizing the importance of blocking both complexes for seizure treatment. Whether these data apply to other mTORopathies due to Tsc1, Tsc2, mTOR, AKT or other gene variants remains to be examined.

Strengths:

The strengths are as follows: 1) they address an important and controversial question that has clinical application, 2) the study uses a reliable and relatively easy method to KO specific genes in cortical neurons, based on AAV9 injections in pups. 2) they perform careful video-EEG analyses correlated with some aspects of cellular pathology.

Weaknesses:

The study has nevertheless a few weaknesses: 1) the conclusions are perhaps a bit overstated. The data do not show that increased mTORC1 or mTORC2 are sufficient to cause epilepsy. However the data clearly show that both increased mTORC1 and mTORC2 activity contribute to the pathology and seizure activity and as such are necessary for seizures to occur. 2) the data related to the EEG would benefit from having more mice. Adding more mice would have helped determine whether there was a decrease in seizure activity with the Rictor or Raptor KO. 3) it would have been interesting to examine the impact of mTORC2 and mTORC1 overexpression related to point #1 above.

Reviewer #3 (Public Review):

Summary: This study investigated the role of mTORC1 and 2 in a mouse model of developmental epilepsy which simulates epilepsy in cortical malformations. Given activation of genes such as PTEN activates TORC1, and this is considered to be excessive in cortical malformations, the authors asked whether inactivating mTORC1 and 2 would ameliorate the seizures and malformation in the mouse model. The work is highly significant because a new mouse model is used where Raptor and Rictor, which regulate mTORC1 and 2 respectively, were inactivated in one hemisphere of the cortex. The work is also significant because the deletion of both Raptor and Rictor improved the epilepsy and malformation. In the mouse model, the seizures were generalized or there were spike-wave discharges (SWD). They also examined the interictal EEG. The malformation was manifested by increased cortical thickness and soma size.

Strengths: The presentation and writing are strong. The quality of data is strong. The data support the conclusions for the most part. The results are significant: Generalized seizures and SWDs were reduced when both Torc1 and 2 were inactivated but not when one was inactivated.

Weaknesses: One of the limitations is that it is not clear whether the area of cortex where Raptor or Rictor were affected was the same in each animal. Also, it is not clear which cortical cells were measured for soma size. Another limitation is that the hippocampus was affected as well as the cortex. One does not know the role of cortex vs. hippocampus. Any discussion about that would be good to add. It would also be useful to know if Raptor and Rictor are in glia, blood vessels, etc.

Author Response

We thank all three Reviewers and the editors for the time and effort they put in reading and critiquing the manuscript. Our revised manuscript includes new data and analyses that address the original concerns. These include, 1) a new Supplemental Figure characterizing Cre expression and cellular phenotypes in the hippocampus, 2) new tables that give a more comprehensive picture of the EEG recordings and statistical analyses, 3) addition of whole cell electrophysiology data, and 4) rewriting to ensure that we do not state that either mTORC1 or mTORC2 hyperactivation is sufficient to cause epilepsy. We discuss the issue of statistical power to detect reduction in generalized seizure rate in the responses below. These suggestions and additions have improved the paper and we hope they will raise both significance and strength of support for the conclusions.

Reviewer #1 (Public Review):

Hyperactivation of mTOR signaling causes epilepsy. It has long been assumed that this occurs through overactivation of mTORC1, since treatment with the mTORC1 inhibitor rapamycin suppresses seizures in multiple animal models. However, the recent finding that genetic inhibition of mTORC1 via Raptor deletion did not stop seizures while inhibition of mTORC2 did, challenged this view (Chen et al, Nat Med, 2019). In the present study, the authors tested whether mTORC1 or mTORC2 inhibition alone was sufficient to block the disease phenotypes in a model of somatic Pten loss-of-function (a negative regulator of mTOR). They found that inactivation of either mTORC1 or mTORC2 alone normalized brain pathology but did not prevent seizures, whereas dual inactivation of mTORC1 and mTORC2 prevented seizures. As the functions of mTORC1 versus mTORC2 in epilepsy remain unclear, this study provides important insight into the roles of mTORC1 and mTORC2 in epilepsy caused by Pten loss and adds to the emerging body of evidence supporting a role for both complexes in the disease development.

Strengths:

The animal models and the experimental design employed in this study allow for a direct comparison between the effects of mTORC1, mTORC2, and mTORC1/mTORC2 inactivation (i.e., same animal background, same strategy and timing of gene inactivation, same brain region, etc.). Additionally, the conclusions on brain epileptic activity are supported by analysis of multiple EEG parameters, including seizure frequencies, sharp wave discharges, interictal spiking, and total power analyses.

Weaknesses:

(1) The sample size of the study is small and does not allow for the assessment of whether mTORC1 or mTORC2 inactivation reduces seizure frequency or incidence. This is a limitation of the study.

We agree that this is a minor limitation of the present study, however, for several reasons we decided not to pursue this question by increasing the number of animals. First, we performed a power analysis of the existing data. This analysis showed that we would need to use 89 animals per group to detect a significant difference (0.8 Power, p= 0.05, Mann-Whitney test) in the frequency of generalized seizures in the Pten-Raptor group and 31 animals per group in the Pten-Rictor group versus Pten alone. It is simply not feasible to perform video-EEG monitoring on this many animals for a single study. Second, even if we did do enough experiments to detect a reduction in seizure frequency, it is clear that neither Rptor nor Rictor deletion provides the kind normalization in brain activity that we seek in a targeted treatment. Both Pten-Rptor and Pten-Rictor animals still have very frequent spike-wave events (Fig. 3D) and highly abnormal interictal EEGs (Fig. 4), suggesting that even if generalized seizures were reduced, epileptic brain activity persists. This is in contrast to the triple KO animals, which have no increase in SWD above control level and very normal interictal EEG.

(2) The authors describe that they inactivated mTORC1 and mTORC2 in a new model of somatic Pten loss-of-function in the cortex. This is slightly misleading since Cre expression was found both in the cortex and the underlying hippocampus, as shown in Figure 1. Throughout the manuscript, they provide supporting histological data from the cortex. However, since Pten loss-of-function in the hippocampus can lead to hippocampal overgrowth and seizures, data showing the impact of the genetic rescue in the hippocampus would further strengthen the claim that neither mTORC1 nor mTORC2 inactivation prevents seizures.

Thank you for pointing out this issue. Cre expression was observed in both the cortex and the dorsal hippocampus in most animals, and we agree that differences in cortical versus hippocampal mTOR signaling could have differential contributions to epilepsy. We initially focused our studies on the cortex because spike-and-wave discharge, the most frequent and fully penetrant EEG phenotype in our model, is associated with cortical dysfunction. In our revised submission we have included a new Figure that quantifies Cre expression in the hippocampal subfields, as well as pS6, pAkt and soma size. These new data show that the amount of Cre expression in the hippocampus is not related to the occurrence of generalized seizures. The pattern of cell size changes in hippocampal neurons is the same as observed in cortical neurons. The levels of pS6 and pAkt are not much changed in the hippocampus, likely due to the sparse Cre expression there. We interpret these findings as supporting the conclusion that the reason we do not see seizure prevention by mTORC1 or mTORC2 inactivation is not due to hippocampal-specific dysfunction.

(3)Some of the methods for the EEG seizure analysis are unclear. The authors describe that for control and Pten-Raptor-Rictor LOF animals, all 10-second epochs in which signal amplitude exceeded 400 μV at two time-points at least 1 second apart were manually reviewed, whereas, for the Pten LOF, Pten-Raptor LOF, and Pten-Rictor LOF animals, at least 100 of the highest- amplitude traces were manually reviewed. Does this mean that not all flagged epochs were reviewed? This could potentially lead to missed seizures.

We reviewed at least 48 hours of data from each animal manually. All seizures that were identified during manual review were also identified by the automated detection program. It is possible but unlikely that there are missed seizures in the remaining data. We have added these details to the Methods of the revised submission.

(4) Additionally, the inclusion of how many consecutive hours were recorded among the ~150 hours of recording per animal would help readers with the interpretation of the data.

Thank you for this recommendation. Our revised submission includes a table with more information about the EEG recordings in the revised version of the manuscript. The number of consecutive hours recorded varied because the wireless system depends on battery life, which was inconsistent, but each animal was recorded for at least 48 consecutive hours on at least two occasions.

(5) Finally, it is surprising that mTORC2 inactivation completely rescued cortical thickness since such pathological phenotypes are thought to be conserved down the mTORC1 pathway. Additional comments on these findings in the Discussion would be interesting and useful to the readers.

We agree that the relationship between mTORC2, cortical thickness, and growth in general is an interesting topic with conflicting results in the literature. We didn’t add anything to the Discussion along these lines because we are up against word limits, but comment here that soma size was increased 120% by Pten inactivation and partially normalized to a 60% increase from Controls by mTORC2 inactivation (Fig. 2C). We and others have previously shown that mTORC2 inactivation (Rictor deletion) in neurons reduces brain size, neuron soma size, and dendritic outgrowth (PMIDs: 36526374, 32125271, 23569215). In our revised submission we also include new data showing that the membrane capacitance of Pten-Ric LOF neurons is normal. Thus, we do not find it completely surprising that mTORC2 inactivation reduces the cortical thickness increase caused by Pten loss. There may still be a slight increase in cortical thickness in Pten-Rictor animals, but it is statistically indistinguishable from Controls.

Reviewer #2 (Public Review):

Summary:

The study by Cullen et al presents intriguing data regarding the contribution of mTOR complex 1 (mTORC1) versus mTORC2 or both in Pten-null-induced macrocephaly and epileptiform activity. The role of mTORC2 in mTORopathies, and in particular Pten loss-off-function (LOF)-induced pathology and seizures, is understudied and controversial. In addition, recent data provided evidence against the role of mTORC1 in PtenLOF-induced seizures. To address these controversies and the contribution of these mTOR complexes in PtenLOF-induced pathology and seizures, the authors injected a AAV9-Cre into the cortex of conditional single, double, and triple transgenic mice at postnatal day 0 to remove Pten, Pten+Raptor or Rictor, and Pten+raptor+rictor. Raptor and Rictor are essentially binding partners of mTORC1 and mTORC2, respectively. One major finding is that despite preventing mild macrocephaly and increased cell size, Raptor knockout (KO, decreased mTORC1 activity) did not prevent the occurrence of seizures and the rate of SWD event, and aggravated seizure duration. Similarly, Rictor KO (decreased mTORC2 activity) partially prevented mild macrocephaly and increased cell size but did not prevent the occurrence of seizures and did not affect seizure duration. However, Rictor KO reduced the rate of SWD events. Finally, the pathology and seizure/SWD activity were fully prevented in the double KO. These data suggest the contribution of both increased mTORC1 and mTORC2 in the pathology and epileptic activity of Pten LOF mice, emphasizing the importance of blocking both complexes for seizure treatment. Whether these data apply to other mTORopathies due to Tsc1, Tsc2, mTOR, AKT or other gene variants remains to be examined.

Strengths:

The strengths are as follows: 1) they address an important and controversial question that has clinical application, 2) the study uses a reliable and relatively easy method to KO specific genes in cortical neurons, based on AAV9 injections in pups. 2) they perform careful video-EEG analyses correlated with some aspects of cellular pathology.

Weaknesses:

The study has nevertheless a few weaknesses: 1) the conclusions are perhaps a bit overstated. The data do not show that increased mTORC1 or mTORC2 are sufficient to cause epilepsy. However the data clearly show that both increased mTORC1 and mTORC2 activity contribute to the pathology and seizure activity and as such are necessary for seizures to occur.

We agree that our findings do not directly show that either mTORC1 or mTORC2 hyperactivity are sufficient to cause seizures, as we do not individually hyperactivate each complex in the absence of any other manipulation. We interpreted our findings in this model as suggesting that either is sufficient based on the result that there is no epileptic activity when both are inactivated, and thus assume that there is not a third, mTOR-independent, mechanism that is contributing to epilepsy in Pten, Pten-Raptor, and Pten-Rictor animals. In addition, the histological data show that Raptor and Rictor loss each normalize activity through mTORC1 and mTORC2 respectively, suggesting that one in the absence of the other is sufficient. However, we agree that there could be other potential mTOR-independent pathways downstream of Pten loss that contribute to epilepsy. We have revised the manuscript to reflect this.

(2) The data related to the EEG would benefit from having more mice. Adding more mice would have helped determine whether there was a decrease in seizure activity with the Rictor or Raptor KO.

Please see response to Reviewer 1’s first Weakness.

(3) It would have been interesting to examine the impact of mTORC2 and mTORC1 overexpression related to point #1 above.

We are not sure that overexpression of individual components of mTORC1 or mTORC2 would result in their hyperactivation or lead to increases in downstream signaling. We believe that cleanly and directly hyperactivating mTORC1 or especially mTORC2 in vivo without affecting the other complex or other potential interacting pathways is a difficult task. Previous studies have used mTOR gain-of-function mutations as a means to selectively activate mTORC1 or pharmacological agents to selectively activate mTORC2, but it not clear to us that the former does not affect mTORC2 activity as well, or that the latter achieves activation of mTORC2 targets other than p-Akt 473, or that it is truly selective. We agree that these would be key experiments to further test the sufficiency hypothesis, but that the amount of work that would be required to perform them is more that what we can do in this Short Report.

Reviewer #3 (Public Review):

Summary: This study investigated the role of mTORC1 and 2 in a mouse model of developmental epilepsy which simulates epilepsy in cortical malformations. Given activation of genes such as PTEN activates TORC1, and this is considered to be excessive in cortical malformations, the authors asked whether inactivating mTORC1 and 2 would ameliorate the seizures and malformation in the mouse model. The work is highly significant because a new mouse model is used where Raptor and Rictor, which regulate mTORC1 and 2 respectively, were inactivated in one hemisphere of the cortex. The work is also significant because the deletion of both Raptor and Rictor improved the epilepsy and malformation. In the mouse model, the seizures were generalized or there were spike-wave discharges (SWD). They also examined the interictal EEG. The malformation was manifested by increased cortical thickness and soma size.

Strengths: The presentation and writing are strong. The quality of data is strong. The data support the conclusions for the most part. The results are significant: Generalized seizures and SWDs were reduced when both Torc1 and 2 were inactivated but not when one was inactivated.

Weaknesses: One of the limitations is that it is not clear whether the area of cortex where Raptor or Rictor were affected was the same in each animal.

Our revised submission includes new data showing that the area of affected cortex and hippocampus are similar across groups. (Figure 1A and Supplementary Figure 1)

Also, it is not clear which cortical cells were measured for soma size.

Soma size was measured by dividing Nissl stain images into a 10 mm2 grid. The somas of all GFP-expressing cells fully within three randomly selected grid squares in Layer II/III were manually traced. Three sections per animal at approximately Bregma -1.6, -2,1, and -2.6 were used. As Cre expression was driven by the hSyn promoter these cells include both excitatory and inhibitory cortical neurons.

Another limitation is that the hippocampus was affected as well as the cortex. One does not know the role of cortex vs. hippocampus. Any discussion about that would be good to add.

See response to Reviewer 1’s second Weakness.

It would also be useful to know if Raptor and Rictor are in glia, blood vessels, etc.

Raptor and Rictor are thought to be ubiquitously active in mammalian cells including glia and endothelial cells. Previous studies have shown that mTOR manipulation can aﬀect astrocyte function and blood vessel organization, however, our study induced gene knockout using an AAV that expressed Cre under control of the hSyn promoter, which has previously been shown to be selective for neurons. Manual assessment of Cre expression compared with DAPI, NeuN, and GFAP stains suggested that only neurons were aﬀected.

Recommendations for the authors: please note that you control which revisions to undertake from the public reviews and recommendations for the authors

Reviewer #1 (Recommendations For The Authors):

In addition to the comments in the public review, it is recommended that the authors provide a more representative ﬁgure for p-Akt staining in the Pten LOF condition in Figure 1 D2. The current ﬁgure is not convincing.

Thanks for the suggestion. We have replaced the images with zoomed in panels that beter demonstrate the diﬀerence.

Additionally, in the last paragraph of the discussion, there is a reference error to an incorrect paper (reference 18) that should be corrected.

Thanks, corrected.

Reviewer #2 (Recommendations For The Authors):

Major comments:

Comment 1: Some statements need clariﬁcations or changes.

(1) Abstract: "spontaneous seizures and epileptiform activity persisted despite mTORC1 or mTORC2 inactivation alone but inactivating both mTORC1 and mTORC2 normalized pathology." Did inactivation of one only also normalized the pathology? Did inactivating both normalized the seizures? Pathology is not equal to seizures.

We have altered this statement to avoid ambiguity.

(2) Abstract: "These results suggest that hyperactivity of both mTORC1 and mTORC2 are suﬃcient to cause epilepsy,". Based on the abstract, it is not clear that it is suﬃcient. It is necessary.

We have altered this statement by removing the term “suﬃcient.”

(3) "Thus, there is strong evidence that hyperactivation of mTORC1 downstream of PTEN disruption causes the macrocephaly, epilepsy, early mortality, and synaptic dysregulation observed in humans and model organisms [17]" I would suggest adding that the strongest evidence is that mTOR GOF mutations lead to the same pathology and epilepsy, suggesting mTORC1 is suﬃcient. The other ﬁndings suggest that it is necessary.

Unless we misunderstand the Reviewer’s point, we believe this viewpoint is already encompassed by the proceeding text that “These phenotypes resemble those observed in models of mTORC1- speciﬁc hyperactivation.”

(4) Introduction (end): "suggesting that hyperactivity of either complex can lead to neuronal hyperexcitability and epilepsy".

Comment 2: I do not agree with the title based on comment 1 above. You did not provide evidence that the mTORCs cause seizures. Your data suggest that they are necessary for seizures or contribute to seizures, but there is no evidence that mTORC2 can induce seizure.

We softened the title by replacing “cause” with “mediate.”

Comment 3: Fig. 1B. Could you beter describe the aﬀected regions. I can see other regions than just the cortex and hippocampus.

Almost all aﬀected cell bodies were in the cortex and hippocampus. The virus in the image is cell-ﬁlling and as such projections from aﬀected neurons throughout the brain can also be seen. We have clariﬁed this in the ﬁgure legend.

Comment 4: I feel unease about the number of animals recorded for EEG to assess seizure frequency. There is not enough power to draw clear conclusions. So, please make sure to not oversell your ﬁndings since it is all-or-nothing data (seizure or no seizure) in this case and the seizure frequency could very well be decreased with single mTOR LOF, but it is impossible to conclude. Maybe discuss this limitation of your study.

We have addressed this in the public comments response.

Minor:

(1) Pten LOF: deﬁne the abbreviation.

Done

(2) Make sure that gene name in mice are not capitalized and italicized.

OK

(3) Fig 1C: could you specify in the results where the analysis was done.

Detail added to Methods (to keep Results concise for word limit)

(4) In the subtitle: "Concurrent mTORC1/2 inactivation, but neither alone, rescues epilepsy and interictal EEG abnormalities in focal Pten LOF". Replace "rescues" but prevents. This is not a rescue experiment since the LOF is done at the same time.

OK

(5) "GS did not appear to be correlated with mTOR pathway activity (Supplementary Figure 2)." Please can you do proper correlation analysis, by plotting all the values as a function of seizure frequency independent of the condition? There is also no correlation between pAKt and seizures.

Here are those data in Author response image 1. They are now part of Supplementary Figure 2.

Reviewer #3 (Recommendations For The Authors):

Figures 1 D, and E show images that are too small to judge. Where are the layers? Please add marks.

We replaced these images with larger zoomed in images to show group differences more clearly. The images no longer show multiple differentiable cortical layers.

If Fig 1 characterizes the model, where is the seizure data? When did they start? Where did they start? Was the focus of the cortical area aﬀected by PTEN loss of function?

Updated figure name to reflect content. Information about the seizure phenotypes is included in Figure 3.

Figure 2 The font size for the calibration is too small. The correlations are hard to see. Colors are not easy to discriminate.

We edited the figure to correct these problems.

Figure 3 shows a clear eﬀect on generalized seizures but the text of the Results does not reﬂect that.

We wanted to be cautious about interpreting these data based on the issue raised by other reviewers that they are underpowered to detect seizure reduction in the Pten-Raptor and Pten-Rictor groups. We have updated the language to atempt to strike a beter balance between over- and under-interpretation. We also performed an additional analysis of the occurrence of generalized seizures to emphasize that only Control and PtRapRic animals have signiﬁcantly lower seizure occurrence that Pten LOF mice (Fig 3C).

For interictal power, was the same behavioral state chosen? Was a particular band aﬀected?

Epochs to be analyzed were selected automatically and were agnostic to behavioral state. Band-speciﬁc eﬀects are outlined in Figure 4B and Table [2].

There is no information about whether the model exhibits altered sleep, food intake, weight, etc.

We didn’t collect information on food intake. It would be possible to look at sleep from the EEG, but that is not something that we are prepared to do at this point. Weight at endpoint was not diﬀerent between genotypes but we did not collect longitudinal data on weight.

Were the sexes diﬀerent?

Included in new Table [1]

Where were EEG electrodes and were they subdural or not?

Additional detail on this has been added to Methods. The screws are placed in the skull but above the dura.

How long were continuous EEG records- the method just says 150 hr. per mouse in total.

Included in new Table [1]

The statistics don't discuss power, normality, whether variance was checked to ensure it did not diﬀer signiﬁcantly between groups, or whether data are mean +- sem or sd. For ANOVAs, were there multifactorial comparisons and what were F, df, and p values? Exact p for post hoc tests?

We have added a new table (Table [3]) that gives information on the exact test used, F, p values, and exact p for post hoc tests. Information regarding power, normality, variance, post- tests and multiple comparisons corrections have been added to Methods section “Statistical Analysis.”

Version 2

Abstract

Abstract

Gene variants that hyperactivate PI3K-mTOR signaling in the brain lead to epilepsy and cortical malformations in humans. Some gene variants associated with these pathologies only hyperactivate mTORC1, but others, such as PTEN, PIK3CA, and AKT, hyperactivate both mTORC1- and mTORC2-dependent signaling. Previous work established a key role for mTORC1 hyperactivity in mTORopathies, however, whether mTORC2 hyperactivity contributes is not clear. To test this, we inactivated mTORC1 and/or mTORC2 downstream of early Pten deletion in a new model of somatic Pten loss-of-function (LOF) in the cortex and hippocampus. Spontaneous seizures and epileptiform activity persisted despite mTORC1 or mTORC2 inactivation alone, but inactivating both mTORC1 and mTORC2 simultaneously normalized brain activity. These results suggest that hyperactivity of both mTORC1 and mTORC2 can cause epilepsy, and that targeted therapies should aim to reduce activity of both complexes.

Introduction

Genetic disruptions that hyperactivate mTOR activity, termed ‘mTORopathies’, cause macrocephaly, intractable childhood epilepsies, and behavioral presentations including autism spectrum disorders [1–4]. The mTOR kinase can function in two complexes, mTORC1 and mTORC2, each with unique downstream targets and upstream activators [5]. A unifying feature of mTORopathies is hyperactivation of mTORC1, which has led to the hypothesis that this complex and its effectors mediate disease phenotypes [6]. A subset of mTORopathies however, (e.g. those caused by variants in PTEN, PIK3CA, and AKT), hyperactivate both mTORC1 and mTORC2 [7,8], and mTORC2 hyperactivity has also been reported in human TLE [9]. It is unclear whether mTORC2 hyperactivity contributes to, or can cause, disease phenotypes such as epilepsy.

Pten LOF in neurons induces soma hypertrophy, increased dendritic branching, and synaptic hyperconnectivity [10,11]. These phenotypes resemble those observed in models of mTORC1-specific hyperactivation [12], and can be prevented with either the mTORC1 inhibitor rapamycin or loss of the mTORC1-specific protein RAPTOR [13]. Rapamycin also rescues epilepsy and mortality associated with Pten loss, even after epilepsy has been established, suggesting that mTORC1 hyperactivity underlies these phenotypes [14–16]. Thus, there is strong evidence that hyperactivation of mTORC1 downstream of PTEN disruption causes the macrocephaly, epilepsy, early mortality, and synaptic dysregulation observed in humans and model organisms [17].

There is also evidence against the mTORC1-centric hypothesis. Rapamycin may affect mTORC2 activity at high doses or over long periods of time [18,17], and mTORC2 regulates synapse function, neuronal size, and cytoskeletal organization independent of mTORC1 [19,20]. In a mouse model in which Pten was inactivated in forebrain neurons in early adulthood, mTORC2 inhibition, but not mTORC1 inhibition, rescued seizures and behavioral abnormalities. In this model, epilepsy was dissociated from macrocephaly, which was rescued by mTORC1 inhibition but not mTORC2 inhibition [21]. However, mTORC2 inactivation did not normalize spontaneous seizures or seizure susceptibility in a model of Pten loss in dentate granule neurons [22]. Thus, there is conflicting evidence about whether mTORC1 or mTORC2 hyperactivity causes epilepsy downstream of Pten loss.

To address this, we suppressed mTORC1 and mTORC2 activity alongside Pten loss by inducing simultaneous deletion of Rptor or Rictor, whose protein products are unique and essential components of mTORC1 and mTORC2, respectively, in a model of somatic Pten LOF. In this model, epilepsy could be rescued by concurrent mTORC1 and mTORC2 inactivation, but persisted when either gene remained intact, suggesting that hyperactivity of either complex can lead to neuronal hyperexcitability and epilepsy.

Results

Generation of a developmental brain somatic mosaic model of Pten LOF. To model the epileptogenic somatic mutations often observed in mTORopathies [23,24], we injected an AAV9 virus expressing GFP and Cre under control of the hSyn promoter into one hemisphere of the cortex of Pten(fl/fl), Pten(fl/fl)-Rptor(fl/fl), Pten(fl/fl)-Rictor(fl/fl), and Pten(fl/fl)-Rptor(fl/fl)-Rictor(fl/fl)mice at P0 (Fig. 1A). We hereafter refer to these mice as Pten LOF, Pten-Rap LOF, Pten-Ric LOF, and PtRapRic LOF. Green fluorescent protein (GFP) was expressed in neurons throughout the cortical layers in all experimental animals, but largely confined to one cortical hemisphere and the underlying hippocampus (Fig. 1B and Supplementary Fig. 1). The percentage of cells that were GFP+ was similar across groups, although there was a decrease in cell density in the Pten LOF and Pten-Ric LOF groups [25] (Fig. 1C).

Histological characterization of the focal cortical Pten LOF model.

A) Experimental timeline showing induction of hSyn-Cre-GFP or Control hSyn-GFP AAV at P0 and EEG recording in adulthood. The location of cortical Cre expression relative to Bregma did not differ between groups. B) Representative images of GFP expression in a Control mouse brain, demonstrating expression predominantly in one hemisphere of the cortex. Projections from affected neurons can be seen in white matter tracts including the fornix, internal capsule, corpus callosum, and cerebral peduncle. C1) Quantification of lesion severity and neuron density in Cre-expressing animals. There were fewer Cre-expressing neurons per unit area in the Pten LOF and Pten-Ric LOF groups, which was at least partially attributable to a decrease in cell density in these groups. No significant differences in Cre expression remained when Cre expression was calculated based on cell density rather than area. C2) Representative images of DAPI and Cre fluorescence in the cortex of Pten LOF and Pten-Rap LOF animals. Cell density and Cre density in Pten-Rap LOF animals was indistinguishable from Controls. D1) Phospho-S6, a marker of mTORC1 activity, was increased by Pten LOF and reduced to control levels by concurrent Rptor loss. Phospho-S6 was also increased from Control levels in Pten-Ric LOF, indicating that mTORC1 hyperactivity was not normalized by Rictor loss. Combined Rptor/Rictor loss also normalized phospho-S6 expression. D2) phospho-Akt, a marker of mTORC2 activity, was increased in Pten LOF and normalized by Rictor loss, but not by Rptor. Combined Rptor/Rictor loss also normalized phospho-Akt expression. Error bars show mean ± s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, *** indicates p<0.001, and **** indicates p<0.0001 as assessed by statistical tests indicated in Table 3. Diagram created with BioRender.com.

We analyzed the impact of Pten, Rptor, and Rictor LOF on mTORC1 and mTORC2 activity in the affected hemisphere using quantitative immunohistochemistry. Phospho-S6 levels, which report mTORC1 activity through S6K1, were increased from Controls in Pten LOF and Pten-Ric LOF brains, but rescued in Pten-Rap LOF brains (Fig. 1D1). Phospho-Akt (S473) levels, which report mTORC2 activity, were increased in Pten LOF and Pten-Rap LOF but normalized in Pten-Ric LOF brains (Fig. 1D2). Levels of both pS6 and pAkt were not different from Controls in PtRapRic LOF brains. These results indicate that Pten LOF hyperactivates both mTORC1 and mTORC2 in this model, and that these increases in activity are independently rescued by mTORC1 and mTORC2 inactivation.

mTORC1 and/or mTORC2 inactivation attenuate Pten LOF-induced macrocephaly, cellular overgrowth, and electrophysiological changes. Macrocephaly and neuronal hypertrophy are well- characterized sequelae of deficient PTEN signaling in patients and mouse models [26,27]. In mouse models of Pten LOF, these phenotypes can be rescued by rapamycin treatment or Rptor LOF [21,28,17,13]. Here, cortical thickness was increased in Pten LOF mice but reduced to Control levels in Pten-Rap LOF, Pten-Ric LOF, and PtRapRic LOF mice (Fig. 2A-B). In individual neurons, Pten LOF showed a ≈100% increase in mean soma size versus Control. Pten-Ric LOF significantly reduced soma size from Pten LOF levels, but still showed a 60% increase over Controls. Soma size in Pten-Rap LOF and PtRapRic LOF neurons did not significantly differ from Controls (Fig. 2C). Whole cell current clamp analysis of neuronal membrane excitability of GFP+ neurons in acute brain slices supported these conclusions, as input resistance, capacitance, and rheobase were all significantly altered by Pten LOF, but not different from control levels in Pten-Rap LOF, Pten-Ric LOF and PtRapRic LOF (Figure 2D). Spontaneous EPSCs were also recorded with whole cell voltage clamp.

Independent mTORC1 or mTORC2 inactivation prevents most cellular effects of Pten LOF, but only dual mTORC1/2 inactivation prevents all.

A) Example images showing fluorescent Nissl stain (magenta) and GFP expression (green) in cortical neurons. The top row shows the cortical thickness in all 5 groups and the bottom row shows zoomed in images depicting the differences in soma size across groups. B) The mean cortical thickness was increased in Pten LOF throughout the cortex. Pten-Rap, Pten-Ric, and PtRapRic LOF cortical thickness did not differ significantly from Controls. C) The mean soma size was strongly increased in Pten LOF and to a smaller extent in Pten-Ric LOF. Pten-Rap LOF and PtRapRic LOF groups did not differ significantly from Controls. D) Whole cell patch clamp analysis of GFP+ neurons showed that capacitance and rheobase were increased in Pten LOF neurons, whereas input resistance was decreased. These changes were not found in in Pten-Rap LOF, Pten-Ric LOF and PtRapRic LOF groups. The frequency of spontaneous EPSCs was elevated in Pten-Rap LOF neurons, and their amplitude was larger in Pten-Ric neurons. Patch data is normalized to values from littermate controls. Error bars show mean ± s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, and **** indicates p<0.0001 as assessed by statistical tests indicated in Table 3.

Pten-Rap LOF and Pten-Ric LOF had elevated sEPSC frequency and amplitude, respectively, as previously reported [22,21,13]. Neither of these parameters were different from Control in PtRapRic LOF neurons (Figure 2D), indicating that only concurrent mTORC1/mTORC2 inactivation can normalize sEPSC parameters.

Concurrent mTORC1/2 inactivation, but neither alone, prevents epilepsy and interictal EEG abnormalities in focal Pten LOF. Next, we measured epileptic brain activity with video-EEG to determine the ability of mTORC1 or mTORC2 activity to rescue this key feature of mTORopathies, Generalized seizures (GS) were observed in 0/9 Control, 4/7 Pten LOF, 2/6 Pten-Rap LOF, 2/7 Pten- Ric LOF, and 0/6 PtRapRic LOF animals, suggesting that only simultaneous mTORC1 and mTORC2 inactivation potentially prevents GS. The frequency of GS events in Pten-Rap LOF and Pten-Ric LOF mice was not significantly lower than the GS frequency in Pten LOF mice, although the number of mice used was not powered to detect reductions in GS frequency. GS events were longer-lasting in Pten-Rap LOF animals than in Pten LOF or Pten-Ric LOF animals (Fig. 3A), and did not appear to be correlated with mTOR pathway activity (Supplementary Fig. 2).

Combined mTORC1 and mTORC2 inactivation, but neither alone, rescues epilepsy in the focal Pten LOF model.

Spontaneous seizures and interictal spike activity were assessed in Pten LOF, Pten-Rap LOF, Pten-Ric LOF, PtRapRic LOF, and Control mice. A) Representative traces of generalized seizures (GS) in a subset of animals in Pten LOF, Pten-Rap LOF, and Pten-Ric LOF groups. B) Spectrograms (top) and traces depicting example SWD event trains. C) Summary data showing the occurrence of GS per animal, number of GS per day, and GS length. GS events were significantly longer in the Pten-Rap LOF group than the other groups. In the Pten LOF group, 19% of GS events (9/48) exceeded 45 seconds in length, and these events were observed in 2/4 Pten LOF GS+ animals. 79% of GS events (15/19) in the Pten-Rap LOF group exceeded 45 seconds, and these events were observed in 2/2 GS+ Pten-Rap LOF animals. 1/11 GS events in the Pten- Ric LOF group exceeded this threshold. D. Summary data showing the SWD rate in all animals. Error bars show mean ±s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, *** indicates p<0.001, and **** indicates p<0.0001 as assessed by tests indicated in Table 3.

The most striking and consistent type of epileptic brain activity in the Pten LOF animals was frequent 5-7 Hz spike trains lasting 3 seconds or more, which fit previous characterizations of spike- and-wave discharges (SWDs) in rodents [29]. SWDs were observed in all Pten LOF, Pten-Rap LOF, and Pten-Ric LOF animals, and in 3/9 Control and 3/6 PtRapRic LOF animals. The frequency of SWDs in Pten LOF, Pten-Rap LOF, and Pten-Ric LOF animals was significantly higher than Controls, but PtRapRic LOF completely blocked this increase. Pten-Ric LOF animals had a lower frequency of SWD events than Pten LOF animals (Fig. 3B). Taken together, these data indicate that inactivating mTORC1 provides no protection against Pten loss-induced epilepsy, and may even exacerbate it. mTORC2 inactivation provides some protection, but only concurrent mTORC1 and mTORC2 inactivation can prevent it.

In addition to epileptic activity, we also found that Pten LOF caused obvious alterations in features of the interictal EEG. To quantify these changes, we measured EEG coastline, absolute mean amplitude, and power spectra of the EEG. Coastline, mean amplitude, and total power were all significantly increased by Pten LOF, and these changes were not significantly reduced by Pten-Rap or Pten-Ric LOF. PtRapRic LOF, however, reduced these increases to Control levels. Correspondingly, EEG power was increased in Pten LOF animals (Fig. 4A; Table 1A). This increase was normalized by PtRapRic LOF, but not by Pten-Rap LOF or Pten-Ric LOF. PtRapRic LOF did not fully rescue a rightward shift in normalized EEG band power (Fig. 4B; Table 1B).

Combined mTORC1 and mTORC2 inactivation, but neither alone, rescues Pten LOF-induced abnormalities in the interictal EEG.

A) Examples of typical EEG traces for each genotype. In EEG epochs that were not characterized as GTCS or SWD events, Pten LOF animals had higher levels of EEG activity as quantified by EEG line length, absolute mean amplitude, and total power. These changes were not significantly decreased in Pten-Rap LOF or Pten-Ric LOF mice, but were normalized in PtRapRic LOF mice. B) Total EEG power was increased by Pten LOF and attenuated, but not normalized, in either Pten-Rap or Pten-Ric LOF mice. Relative power was decreased in delta and increased in higher frequencies by Pten LOF. Pten-Rap LOF, Pten-Ric LOF, and PtRapRic LOF animals all showed a milder rightward shift of EEG power. C) Line length, mean amplitude, and power are increased in Pten LOF and normalized by PtRapRic LOF. Error bars show mean ± s.e.m. ns indicates p>0.05, * indicates p<0.05, ** indicates p<0.01, *** indicates p<0.001 as assessed by statistical tests indicated in Table 3. Two-way ANOVA p-values for EEG power are reported in Table 2.

Discussion

Convergent discoveries have positioned mTORopathies as prime candidates for a precision medicine approach [30,31]. There has been some clinical success with rapamycin analogues [32], but preclinical animal studies have suggested ways to further improve the selectivity and efficacy of the targets [12]. For the subset of mTORopathies in which mTORC1 but not mTORC2 signaling is hyperactive, selective inhibition of mTORC1 and its downstream effectors rescues many animal phenotypes, including seizures [33,34]. Studies using rapamycin as an mTORC1 inhibitor also came to this conclusion in Pten LOF models, but potential inhibition of mTORC2 and the finding that genetic inhibition of mTORC1 via Rptor deletion did not stop seizures, whereas inhibition of mTORC2 did [21], challenged this view. Here, we tested whether mTORC1 or mTORC2 inhibition alone were sufficient to block disease phenotypes in a model of somatic Pten LOF. Neither was, which agrees with previous findings that seizures persist after Rptor LOF and Rictor LOF in separate model systems [21,22]. This data could be interpreted as suggesting that epilepsy downstream of Pten LOF can proceed via mTORC-independent mechanisms, such as elevated ß-catenin or its protein phosphatase activity [25,35]. However, we found that simultaneous mTORC1 and mTORC2 inhibition corrected all of the phenotypes of our model except for a rightward shift in relative EEG power (Fig. 4B). This suggests that epilepsy can be caused by hyperactivity of either mTORC1 or mTORC2-dependent signaling downstream of Pten LOF, but not alternative pathways. Future studies should determine whether this is also true for other gene variants that hyperactivate both mTORC1 and mTORC2, as well in other models that have fully penetrant generalized seizures.

Epilepsy in Pten-Ric LOF mice likely emerges via mechanisms similar to those in mTOR pathway variants such as TSC that hyperactivate mTORC1 but not mTORC2, although additional effects of mTORC2 inactivation may contribute [22]. Epilepsy in Pten-Rap LOF occurs in the absence of macrocephaly, a hallmark of mTORC1 hyperactivity, possibly due to mTORC2 hyperactivity’s effect on synaptic transmission [20]. Roles for both complexes in emergent neural network function are corroborated by findings that behavior can be altered by either mTORC1 or mTORC2 inactivation [36]. Rapamycin derivatives are somewhat selective for mTORC1 [37,38] and are currently being tested in the clinic to treat mTORopathies and PTEN-related disorders [39]. Even more selective targeting of mTORC1 than can be achieved through rapamycin and its derivatives has been proposed as a way to effectively treat disease with fewer side effects [40]. Our data suggest that this may only be true for mTORopathies that selectively hyperactive mTORC1. For others, including PTEN, PIK3CA, and AKT, dual inhibitors or inhibitors of PI3K, which are being tested preclinically [41–43], are likely to be required.

Materials and Methods

Animals

Rptor (Jackson Labs #013188) and Rictor (Jackson Labs #020649) homozygous floxed mice were crossed to a Pten homozygous floxed line ([44] Jackson Labs #006440).

Experimental animals were the offspring of two animals homozygous floxed for Pten and heterozygous floxed for Raptor and/or Rictor. Intracranial viral injections were done at P0 or P1. Animals that were homozygous for the floxed Pten allele and either homozygous or wild-type at the Raptor and/or Rictor allele were injected with an AAV9 viral vector expressing an eGFP-Cre fusion driven by the hSyn promoter (Addgene #105540). Control animals were of the same genotype but were injected with an AAV9 virus expressing eGFP under control of the hSyn promoter, but not Cre (Addgene #105539). Pup injections were conducted with hypothermic anesthesia. A Nanoject (Drummond) was used to deliver 100 nL of Cre or control virus to three sites spanning the left hemisphere of the cortex. Pups were quickly rewarmed and returned to their home cage. Additional control animals with no floxed genes on a C57B6/J background (Jackson Labs #000644) were injected with the eGFP-Cre expressing AAV9 to ensure that Cre expression did not have effects independent of the Ptenfl/flgenotype (Supplementary Figure 3).

Surgeries and EEG Recordings

Animals were aged to six weeks and then implanted with a wireless EEG monitoring headcap. Animals were anesthetized with 4% isofluorane and arranged on a stereotaxic surgical apparatus with ear bars. Anesthesia was maintained with 1.5-2.5% isofluorane. The skull surface was exposed and holes were made with a 23g needle at six locations on the skull surface. 3/32” screws (Antrin Miniature Specialties) were inserted into the skull to record epidural EEG. Screws were secured with VetBond then covered with surgical cement. Recording leads were placed bilaterally at approximately Bregma AP +1.0 ML 1.0, AP -1.0 ML 1.5, and AP -2.5 ML 1.5. The screws were attached to a six-pin Millimax strip which was secured to the skull with additional cement. Animals were administered 5 mg/kg ketoprofen post-surgery and allowed to recover for at least 5 days before EEG recording. Surgical protocol was based on guidance provided by Pinnacle Technologies (https://www.pinnaclet.com).

After recovery from surgery, animals were fitted with wireless three-channel EEG preamplifiers (Pinnacle Technologies) and EEG signal was recorded at 256 Hz with simultaneous video recording. EEG recording session length was dependent on battery life. A total of at least 150 hours of EEG data was collected for each experimental animal (Table 1).

Histology and imaging

EEG-implanted animals were euthanized by isofluorane overdose followed by transcardiac perfusion with 4% paraformaldehyde. Brains were postfixed in 4% paraformaldehyde for 24 hours and then transferred to 30% sucrose for at least 48 hours. 40 µm coronal slices were cut and preserved in a cryoprotectant solution (50% PBS, 30% ethylene glycol, 20% glycerol). Prior to staining, slices were washed three times with PBS. Eight unstained slices spanning the affected area (Bregma -3.6 through Bregma +1.1) were mounted with DAPI Fluoromount-G (SouthernBiotech) and used to assess cortical thickness, GFP expression, and cell density. GFP expression, cell density, and soma size were measured in the upper layers of the cortex to reduce within-subject variability. GFP expression was also measured in the dentate gyrus, CA1, and CA3 regions of the dorsal hippocampus. For other analyses, three sections per animal at approximately Bregma -1.6, -2,1, and -2.6 were used. A fluorescent Nissl stain (Invitrogen N21482, 1:50) was used to assess soma size.

After washing, slices were placed in blocking solution (10% normal goat serum, 0.1% Triton X-100, and PBS) for 1 hour. Slices were incubated in the following primary antibodies for 3 hours at room temperature: phospho-S6 Ribosomal Protein Ser240/244 (rabbit monoclonal, 1:1000 dilution, Cell Signaling Technology, catalog #5364, RRID: AB_10694233), phospho-AKT Ser473 (rabbit monoclonal, 1:1000 dilution, Cell Signaling Technology, catalog #4060), and NeuN (guinea pig polyclonal, 1:1000 dilution, Synaptic Systems, catalog #266004). Following primary antibody application, slices were washed three times in PBS and the incubated in AlexaFluor secondary antibodies goat anti-guinea pig 647 and goat anti-rabbit 594 (Invitrogen) for 1 hour at room temperature.

Slides stained for Nissl substance, NeuN/pS6, and NeuN/pAkt were imaged at 10x (2048x2048) with 5 µm z-step on a Nikon C2 confocal microscope (UVM Microscopy Imaging Core). One image of the left (GFP-expressing) cortical hemisphere in each of three region-matched slices per animal was collected for analysis. Widefield images (2560x2160) of unstained DAPI-mounted cross- sectional slices were taken with a Zyla sCMOS camera (Andor) mounted on an upright microscope (IX73; Olympus) at 5x and 20x resolution. Pixel width was manually calculated using an 0.1 mm hemocytometer.

Slice electrophysiology

Slice electrophysiology was conducted at P14-30. Animals were deeply anesthetized and decapitated, and the brain was quickly dissected into ice-cold cutting solution (126 mM NaCl, 25 mM NaHCO3, 10 mM d-glucose, 3.5 mM KCl,1.5 mM NaH2PO4, 0.5 mM CaCl2, 10.0 mM MgCl2, pH 7.3-7.4). 350-um slices were cut using a Leica 1000S Vibratome. Slices were transferred to 37°C aCSF (126 mM NaCl, 3.5 mM KCl, 1.0 mM MgCl2, 2.0 mM CaCl2, 1.5 mM NaH2PO4,25 mM NaHCO3, and 10 mM d-glucose, pH 7.3-7.4) and incubated for 30 min. The slices were then incubated at room temperature for at least another 30 min prior to recording. All solutions were continuously bubbled with 95% O2 and 5% CO2. Individual slices were transferred to a recording chamber located on an upright microscope (BX51; Olympus) and were perfused with heated (32-34°C), oxygenated aCSF (2 mL/min).

Whole-cell voltage-clamp and current-clamp recordings were obtained using Multiclamp 700B and Clampex 10.7 software (Molecular Devices). GFP+ cells in cortical layer 2/3 in the motor cortex were targeted for patching. Intracellular solution contained (in mM): 136 mM K-gluconate, 17.8 mM HEPES, 1 mM EGTA, 0.6 mM MgCl2, 4 mM ATP, 0.3 mM GTP, 12 mM creatine phosphate, and 50 U/ml phosphocreatine kinase, pH 7.2. When patch electrodes were filled with intracellular solution, their resistance ranged from 4–6 MΩ. Access resistance was monitored continuously for each cell.

For current-clamp experiments, the intrinsic electrophysiological properties of neurons were tested by injecting 500-ms square current pulses incrementing in 20 pA steps, starting with -100 pA. The membrane time constant was calculated from an exponential fit of current stimulus offset. Input resistance was calculated from the steady state of the voltage responses to the hyperpolarizing current steps. Membrane capacitance was calculated by dividing the time constant by the input resistance.

Action potentials (APs) were evoked with 0.5 s, 20 pA depolarizing current steps. Rheobase was defined as the minimum current required to evoke an AP during the 500 ms of sustained somatic current injection. For voltage-clamp experiments to measure spontaneous EPSC frequency and amplitude, neurons were held at -70 mV and recorded for 2 min. All electrophysiology data were analyzed offline with AxoGraph X software (AxoGraph Scientific).

EEG analysis

EEG files were converted from Pinnacle’s proprietary format to European Data Format (EDF) files and imported to Matlab. A filtering program was used to flag traces in which seizures were likely to be occurring based on amplitude and line length between data points. For Control and PtRapRic LOF animals, all 10 second epochs in which signal amplitude exceeded 400 µV at two time points at least 1 second apart were manually reviewed by a rater blinded to genotype. This method was not suitable for the Pten LOF, Pten-Raptor LOF, and Pten-Rictor LOF animals because of the density of high-amplitude interictal activity they displayed. For these animals, at least 100 of the highest-amplitude traces were manually reviewed and then traces with persistent abnormally low amplitude, often indicating postictal suppression, were reviewed as well. Flagged traces were displayed for a rater to mark the beginning and end of each seizure. We also reviewed at least 48 hours of data from each animal manually. All seizures that were identified during manual review were also identified by the automated detection program. Spike-and-wave discharge events were manually marked in 8 evenly spaced one-hour epochs in each of two 24-hour recording sessions per animal and verified on video not to be caused by movement such as chewing or scratching.

Baseline EEG measures were taken from a representative sample of EEG files for each animal.

692 evenly spaced five-second epochs were sampled over 24 hours, repeated for two recording sessions per animal. Line length was defined as the sum of the linear distances between adjacent data points during the five-second analysis epoch. Mean amplitude was defined as the mean of the absolute value of data points in the five-second analysis epoch. Power spectral density was calculated with the pwelch function in Matlab. All three EEG channels were analyzed and the mean of all channels was used for statistical analysis.

Image Analysis

Image analysis was conducted using ImageJ/Fiji. Cell density and Cre expression in the cortex were automatically assessed using the Fiji Analyze Particles tool. Cre expression in the hippocampus was manually assessed by counting Cre-expressing punctae within a 100 µm linear portion of the region of interest. Because neuronal somas within the hippocampal cell layers are so closely packed together, we were unable to resolve DAPI punctae to assess cell density in the hippocampus. Soma size was measured by dividing Nissl stain images into a 10 mm2 grid. The somas of all GFP-expressing cells fully within three randomly selected grid squares in Layer II/III were manually traced. pS6 and pAkt expression were measured by drawing 200 µm wide columns spanning all cortical layers. Background was subtracted from images with a 30 µm rolling ball algorithm (Fiji/ImageJ). The mean pixel value of the column was recorded and values were averaged by animal.

Statistical analysis

Prism 9 or 10 (GraphPad Prism) was used to conduct statistical analyses and create graphs. All data are presented as mean ± SEM. Power analyses on preliminary and published data were used to calculate the number of animals necessary for this study, with the exception of GS occurrence, which is discussed below. All data distributions were assessed for normality using the Shapiro–Wilk test. If data did not meet the criteria for normal distribution (p<0.05), Kruskal-Wallis tests were used to assess statistical relationships with Dunn’s post-hoc correction for multiple comparisons. Variance was assessed using Brown-Forsythe tests. If variance differed significantly (p<0.05), Welch’s ANOVA test was used to assess statistical relationships.

Group differences in normally distributed datasets with equal variances were assessed using one-way ANOVA with Tukey post-hoc correction for multiple comparisons. Power spectral density was assessed using two-way ANOVA with Tukey post-hoc analysis. Seizure occurrence was assessed with a Generalized Linear Model with a binary distribution and logistic link function implemented in SPSS and Bonferroni corrected for multiple comparisons. See Tables 2 and 3 for details regarding statistical tests. A power analysis showed that an excessive number of animals would be required to detect a decrease in GS incidence or frequency by Rptor or Rictor LOF in this model, thus we chose not to pursue the possibility that Rptor or Rictor loss decreases these outcome measures.